---
title: "Morphological profiling workflows"
author: "Cytomining Hackathon attendees"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphological profiling workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(magrittr)
library(dplyr)
futile.logger::flog.threshold(futile.logger::WARN)
```


## Load data
First, load the data, which is stored in a database backend 

```{r}
fixture <- '~/tmp/BBBC021.sqlite'

if (!file.exists(fixture)) {
  download.file("http://www.broadinstitute.org/~shsingh/BBBC021.sqlite", 
              destfile = fixture, 
              method = "curl", 
              quiet = FALSE, mode = "w",
              cacheOK = TRUE,
              extra = getOption("download.file.extra"))
}

db <- src_sqlite(path = fixture)
```


```{r}
images <-
  tbl(src = db, "supplement_Image") 

objects <- 
  tbl(src = db, "supplement_Object") 

metadata <- 
  tbl(src = db, "supplement_GroundTruth") %>%
  rename(Image_Metadata_Compound = compound,
         Image_Metadata_Concentration = concentration,
         Image_Metadata_MOA = moa) 

images %<>%
  inner_join(
    metadata,
    by = c("Image_Metadata_Compound", "Image_Metadata_Concentration")
  )

data <-
  inner_join(images,
             objects,
             by = c("TableNumber", "ImageNumber")
  ) %>%
  compute()

```


How many rows does this table have?

```{r}
data %>%
  dplyr::tally() %>%
  knitr::kable()
```

All the code after this belongs to a different dataset but can provide some 
clues on how to do the analysis

```{r}
qc_cols <- c("q_debris")

group_cols <-
  c("g_plate",
    "g_well",
    "g_image",
    "g_pattern",
    "g_channel")

feature_cols <-
  colnames(measurements) %>%
  stringr::str_subset("^m_")

measurements %<>%
  dplyr::select(one_of(c(group_cols, qc_cols, feature_cols)))

```

## Clean

Let's remove cells that come from images that were marked as having debris

```{r}
debris_removed <-
  measurements %>% dplyr::filter(q_debris == 0)
```

Then, remove cells where all the measurements are NA's (TODO: explain why this
may be needed)

```{r}
na_rows_removed <-
  cytominr::drop_na_rows(
    population = debris_removed,
    variables = feature_cols
  ) %>%
  dplyr::compute()
```

### Normalize 

We need to normalize the data so that

- features are on the same scale

- plate-to-plate variation is reduced

The default for doing this is `standardization`. Here, we take all the cells
from control wells in the experiment (this is where the external metadata gets
used) and compute normalizations parameters from that (in this case, just the
mean and s.d.) and then apply it to the whole dataset (i.e. the population)

```{r}
normalized <-
  cytominr::normalize(
    population = na_rows_removed,
    variables = feature_cols,
    strata =  c("g_plate", "g_pattern", "g_channel"),
    sample =
      na_rows_removed %>%
      dplyr::inner_join(
        ext_metadata %>% dplyr::filter(Type == "ctrl") %>%
          dplyr::select(g_well)
      )
  )
```

In some cases, we may have features that have no variance at all (e.g. Euler 
number). If these features have not already been removed by this stage, the 
standardization step will results in all values for that feature being NA (
because s.d. = 0). Lets remove them:

First, count how many cells have NA values per feature:

```{r}
na_frequency <-
  cytominr::count_na_rows(
    population = normalized,
    variables = feature_cols)

na_frequency %>%
  tidyr::gather(feature, na_count) %>%
  knitr::kable()

```

As it turns out, no feature has NA in this example. 
But lets run this cleaning operation  anyway (no features will be dropped)

```{r}

cleaned <-
  cytominr::select(
    population = normalized,
    variables = feature_cols,
    operation = "drop_na_columns"
)
```

## Transform

Tranform the data so that assumptions we may later make about the data
distribution are satisfied (e.g. Gaussianity). The default here is 
`generalized_log`. (TODO: explain this further) 

```{r}
transformed <-
  cytominr::transform(
    population = cleaned,
    variables = feature_cols
  )
```

## Select features

Finally, we typically perform feature selection on the data (TODO: explain 
further). Feature selection is an expensive operation, so we usually want to 
train the feature selection model on a sample of the dataset. Here, we choose
to aggregate the data instead of sampling it (i.e. collapse it to per-well
aggregates)

```{r}
aggregated <-
  cytominr::aggregate(
    population = transformed,
    variables = feature_cols,
    strata = group_cols
  ) %>%
  dplyr::collect()
```

... and then apply feature selection on the per-cell data. 
```{r}
selected <-
  cytominr::select(
    population = transformed,
    variables = feature_cols,
    sample = aggregated,
    operation = "correlation_threshold"
  ) %>%
  dplyr::collect()
```

And now lets take a glimpse at the data!
```{r}
selected %>%
  dplyr::glimpse()
```

